\documentclass[11pt,a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}

\geometry{margin=1in}

\title{\textbf{HEP Agent Simulation Framework}\\ \large Model Documentation}
%\author{Development Team}
\date{\today}

\begin{document}

\maketitle
\tableofcontents

\chapter{Introduction}
This document describes the mathematical and logical model underlying the Human Ecology Project (HEP) Agent Simulation.

\chapter{Environment Model}
\section{Grid Structure}
Description of the spatial grid, resolution, and coordinate system.

\section{Habitat Suitability (HEP)}
Description of the HEP index and how it influences agent survival and movement.

\subsection{Population Pressure Scaling (Weibull Equation)}
When population pressure is enabled (\texttt{with\_pop\_pressure = .true.}), the actual available habitat suitability in a cell, $hep_{av}$, is dynamically scaled based on the local human density in that cell compared to its carrying capacity:
\begin{equation}
    hep_{av} = PP \cdot hep
\end{equation}
Where:
\begin{itemize}
    \item $hep$: The baseline habitat suitability read from NetCDF inputs.
    \item $PP$: The population pressure multiplier, modeled using a normalized Weibull probability density function:
\end{itemize}

\begin{equation}
    PP = \frac{ \left( \frac{\eta}{\epsilon} \right) \cdot \left( \frac{\delta_{\rho}}{\epsilon} \right)^{\eta - 1} \cdot \exp\left( - \left( \frac{\delta_{\rho}}{\epsilon} \right)^{\eta} \right) }{ max\_pp }
\end{equation}
With the following definitions:
\begin{itemize}
    \item $\delta_{\rho} = \frac{\rho}{\rho_c}$: The occupancy ratio.
    \item $\rho$: Current smoothed human density in the cell (People / 100 km$^2$).
    \item $\rho_c = N_{max} \cdot hep$: The local carrying capacity of the cell.
    \item $N_{max}$ (\texttt{rho\_max}): Configured carrying capacity density parameter.
    \item $\eta$ (\texttt{eta}): Weibull shape parameter.
    \item $\epsilon$ (\texttt{epsilon}): Weibull scale parameter.
    \item $max\_pp$: Normalization factor representing the maximum value of the Weibull PDF to ensure $PP \le 1.0$:
\end{itemize}

\begin{equation}
    max\_pp = \left( \frac{\eta}{\epsilon} \right) \cdot \left( 1 - \frac{1}{\eta} \right)^{1 - 1/\eta} \cdot \exp\left( - \left( 1 - \frac{1}{\eta} \right) \right)
\end{equation}

\subsection{Parameter Sensitivities to Reduce Suitability Reduction}
To reduce the suitability penalty (i.e., keep $PP$ closer to $1.0$), you can modify the configuration variables in the \texttt{Available Hep} section:
\begin{itemize}
    \item \textbf{Reduce overpopulation penalty}: Increase $N_{max}$ (\texttt{rho\_max}) or $\epsilon$ (\texttt{epsilon}). This increases the capacity threshold $\rho_c$ or broadens the curve so that higher human densities do not trigger the exponential suitability decay.
    \item \textbf{Reduce underpopulation penalty}: Decrease $\eta$ (\texttt{eta}) towards $1.0$. Reducing $\eta$ to exactly $1.0$ completely eliminates the low-density penalty, meaning that completely unpopulated cells ($\rho = 0$) receive $PP = 1.0$.
\end{itemize}

\chapter{Agent Dynamics}
\section{Population Structure}
Agents are defined by a set of intrinsic and extrinsic attributes:
\begin{itemize}
    \item \textbf{Age}: Tracked in ticks (1 tick $\approx$ 1 week).
    \item \textbf{Gender}: Binary classification (Male/Female).
    \item \textbf{Resources}: Energy levels accumulated from grid cells.
\end{itemize}

\section{Movement Rules}
Logic governing agent migration and dispersal.

\section{Demographics}
The demographics model manages population growth and mortality through a combination of individual agent rules and macroscopic feedback loops.

\subsection{Birth Model (Verhulst Pressure)}
To ensure population stability and prevent exponential runaway, the simulation utilizes a \textbf{macroscopic fertility scaling factor} (\texttt{K\_fertility}). 

The target birth rate $\phi_{target}$ is calculated using the Verhulst logistic equation:
\begin{equation}
    \phi_{target} = r \cdot \left( 1 - \frac{N}{N_C} \right)
\end{equation}
Where:
\begin{itemize}
    \item $r$: Intrinsic growth rate.
    \item $N$: Current total population count.
    \item $N_C$: Carrying capacity of the environment.
\end{itemize}

The actual fertility of each female agent is then scaled by $K_{fertility}$ to match this target, bounded by \texttt{Kmin} and \texttt{Kmax} to prevent demographic collapse or excessive volatility.

\subsection{Death Model}
Mortality is driven by multiple factors:
\begin{itemize}
    \item \textbf{Natural Mortality}: Age-dependent risk.
    \item \textbf{Starvation}: Triggered when an agent's resources fall below a critical threshold.
    \item \textbf{Environmental Risk}: Probability of death based on HEP suitability (e.g., drowning in water cells).
\end{itemize}

\chapter{Interaction Model}
\section{Resource Competition}
How agents compete for limited resources in a grid cell.

\chapter{Recent Changes}
\input{changelog}

\end{document}
